---
title: "ggplot aesthetics"
author: "M. Toomey"
date: "`r Sys.Date()`"
output: 
html_document: 
    highlight: tango
    theme: united
editor_options: 
  chunk_output_type: console
---

In addition to mapping data points to Cartesian coordinates in various geometries, `ggplot` allows you to map both continuous and discrete variables to the aesthetic of the geometric objects including: color, shape, and size. Therefore, you can represent the intersection of many different variables in a single plot.  

### Basic mapping of aesthetic variables

#### Packages
```{r}
library(ggplot2)
library(ggthemes)
library(patchwork)
```

Here we will specify the mappings within the `aes()` function and they will be applied to all of the geoms that follow it. 

```{r}
# mapping of a discrete variable to point color
p1 <- ggplot(data=mpg) +                                   
      aes(x=hwy,y=cty,color=class) + 
      geom_point(size=3) 
p1
             
# mapping of a discrete variable to point shape (<= 6)
p1 <- ggplot(data=mpg) +
      aes(x=hwy,y=cty,shape=class) + 
      geom_point(size=3) 
p1

# mapping of a discrete variable to point size (not advised)
p1 <- ggplot(data=mpg) +                                   
      aes(x=hwy,y=cty,size=class) + 
      geom_point() 
p1

# mapping a continuous variable to point size 
p1 <- ggplot(data=mpg) + 
      aes(x=hwy,y=cty,size=displ) + 
      geom_point() 
p1

# mapping a continuous variable to point color
p1 <- ggplot(data=mpg) + 
      aes(x=hwy,y=cty,color=displ) + 
      geom_point(size=5) 
p1

# mapping two variables to different aesthetics
p1 <- ggplot(data=mpg) +                                   
      aes(x=hwy,y=cty,shape=class,color=displ) + 
      geom_point(size=5) 
p1

# use shape for smaller number of categories
p1 <- ggplot(data=mpg) +                                  
      aes(x=hwy,y=cty,shape=drv,color=fl) + 
      geom_point(size=5) 

# use all 3 (size, shape, color) to indicate 5 attributes!
p1 <- ggplot(data=mpg) +                                   
      aes(x=hwy,
          y=cty,shape=drv,
          color=fl,
          size=displ) + 
      geom_point()
p1

# mapping a variable to the same aesthetic in two different geoms
p1 <- ggplot(data=mpg) + 
            aes(x=hwy,y=cty,color=drv) + 
            geom_point(size=2) +                                 
            geom_smooth(method="lm")
p1

# break out the drive types (note what fill affects)
p1 <- ggplot(data=mpg) + 
            aes(x=hwy,y=cty, fill=drv) +
            geom_point(size=2) +                                 
            geom_smooth(method="lm")
p1


# use both if you want points, lines, and confidence intervals colored
p1 <- ggplot(data=mpg) + 
            aes(x=hwy,y=cty, color=drv, fill=drv) +
            geom_point(size=2) +                                 
            geom_smooth(method="lm")
p1

# group allows for grouping by a variable without specify a color, size, shape. Here the grouping only affects how the fit line is plotted
p1 <- ggplot(data=mpg) +
          aes(x=hwy,y=cty, group=drv) +
          geom_point(size=2) +                                 
          geom_smooth(method="lm")
p1


```

#### Mapping aesthetics within geoms

When plotting multiple geoms you may want to map the aesthetics differently within these. This can be done by specify the mapping with the geom function. 

```{r, warning=FALSE, message=FALSE}

# now use aesthetic mappings within each geom to over-ride defaults
# subset the data frame to pull out what you need

p1 <- ggplot(data=mpg) + 
      aes(x=hwy,y=cty,col=drv) +
      geom_point(data=mpg[mpg$drv=="4",]) + 
      geom_smooth()
p1

# instead of subsetting, just map an aesthetic
p1 <- ggplot(data=mpg) + 
      aes(x=hwy,y=cty) +
      geom_point(aes(color=drv)) + 
      geom_smooth()
p1

# Conversely, map the smoother, but not the points
p1 <- ggplot(data=mpg) + 
      aes(x=hwy,y=cty) +
      geom_point() + 
      geom_smooth(aes(color=drv))
p1

# also, subset in the first layer to eliminate some data entirely
# instead of subsetting, just map an aesthetic
p1 <- ggplot(data=mpg[mpg$drv!="4",]) + 
      aes(x=hwy,y=cty) +
      geom_point(aes(color=drv)) + 
      geom_smooth()
p1

```

```
Lesson developed with materials adapted from: https://gotellilab.github.io/Bio381/
```