---
title: "Arranging multiple plots with facet and patchwork"
author: "M. Toomey"
date: "`r Sys.Date()`"
output: 
html_document: 
    highlight: tango
    theme: united
editor_options: 
  chunk_output_type: console
---

Multi-panel figures are a common feature of publications in many disciplines and a useful way to display complementary data plots. However, combining and arranging separate plots in a precisely align and repeatable way can be difficult with typical tools. The package [patchwork](https://patchwork.data-imaginist.com/) works with `ggplot` and provides an intuitive way to combine plots. 

#### Packages
```{r}
library(ggplot2)
library(ggthemes)
library(patchwork)
```

###lotting multiple panel graphs with `patchwork`
```{r,message=FALSE,warning=FALSE}
g1 <- ggplot(data=mpg) +
        aes(x=hwy,y=cty) + 
        geom_point() + 
        geom_smooth()
g1

g2 <- ggplot(data=mpg) +
        aes(x=fl,fill=I("maroon4"),color=I("black")) +
        geom_bar(stat="count") + 
        theme(legend.position="none")
g2

g3 <- ggplot(data=mpg) +
        aes(x=hwy,fill=I("slateblue"),color=I("black")) + 
        geom_histogram()
g3
 
  g4 <- ggplot(data=mpg) +
        aes(x=fl,y=cty,fill=fl) + 
        geom_boxplot() + 
        theme(legend.position="none")
  
g4

# place two plots horizontally
  g1 + g2

# place 3 plots vertically
  g1 + g2 + g3 + plot_layout(ncol=1)
  
# change relative area of each plot
  g1 + g2 + plot_layout(ncol=1,heights=c(2,1))
  
  g1 + g2 + plot_layout(ncol=2,widths=c(1,2))
  
# add a spacer plot (under construction)
  g1 + plot_spacer() + g2
  
# use nested layouts
  g1 + {
    g2 + {
      g3 +
        g4 +
        plot_layout(ncol=1)
    }
  } +
    plot_layout(ncol=1)
  
# - operator for subtrack placement
  g1 + g2 - g3 + plot_layout(ncol=1)
  
# / and | for intuitive layouts
  (g1 | g2 | g3)/g4
  
  (g1 | g2)/(g3 | g4)
  
  # Add title, etc. to a patchwork
g1 + g2 + plot_annotation('This is a title', caption = 'made with patchwork')

# Change styling of patchwork elements
g1 + g2 +
  plot_annotation(
    title = 'This is a title',
    caption = 'made with patchwork',
    theme = theme(plot.title = element_text(size = 16))
  )

# Add tags to plots
g1 / (g2 | g3) +
  plot_annotation(tag_levels = 'a')
```

### Swapping axis orientation for multipanel plot

```{r, message=FALSE, warning=FALSE}
g3a <- g3 + scale_x_reverse()
g3b <- g3 + scale_y_reverse() 
g3c <- g3 + scale_x_reverse() + scale_y_reverse()

(g3 | g3a)/(g3b | g3c)
(g3 + coord_flip() | g3a + coord_flip())/(g3b + coord_flip() | g3c + coord_flip())
```

### Faceting variables to generate multiple plots

The faceting functions of ggplot allow you to separate and plot subsets of your data by variables within the data set. 

```{r,warning=FALSE,message=FALSE}
# basic faceting with variables split by row, column, or both
m1 <- ggplot(data=mpg) + 
            aes(x=hwy,y=cty) + 
            geom_point() 

m1 +  facet_grid(class~fl)     
m1 + facet_grid(class~fl, scales="free_y")
m1 + facet_grid(class~fl, scales="free")
m1 + facet_grid(.~class)
m1 + facet_grid(class~.)

# use facet wrap when variables are not crossed
m1 + facet_grid(.~class)
m1 +  facet_wrap(~class)     
m1 + facet_wrap(~class + fl)
m1 + facet_wrap(~class + fl, drop=FALSE)
m1 + facet_grid(class~fl)

# use facet with other aesthetic mapping within rows or columns

m1 <- ggplot(data=mpg) +                                         
        aes(x=displ,y=cty,color=drv) + 
        geom_point() 

m1 + facet_grid(.~class)

# easy to switch to other geoms
m1 <- ggplot(data=mpg) +                                        
        aes(x=displ,y=hwy,color=drv) + 
        geom_smooth(se=FALSE,method="lm") 

m1 + facet_grid(.~class)

# fitting with boxplots over a continuous variable
m1 <- ggplot(data=mpg) +  
        aes(x=hwy,y=cty) + 
        geom_boxplot()

m1 + facet_grid(.~class)

# add a group and fill mapping for subgroups
m1 <- ggplot(data=mpg) +                                        
        aes(x=displ,y=hwy,group=drv,fill=drv) + 
        geom_boxplot()

m1 + facet_grid(.~class)

```

```
Lesson developed with materials adapted from: https://gotellilab.github.io/Bio381/
```