---
title: "Data Import and Export with `readr`"
author: "M. Toomey"
date: "`r Sys.Date()`"
output: 
html_document: 
    highlight: tango
    theme: united
editor_options: 
  chunk_output_type: console
---

Readr is a package that is part of Tidyverse with a range of functions to facilitate the parsing and import of data files into R. It replicates and expands many of the existing function in base R

### Importing .csv files

Download my ebird dataset from my github account here:  https://github.com/mbtoomey/Biol_7263/blob/main/Data/MBT_ebird.csv?raw=true

Create a "Data" folder in your project and save this files there. 

```{r eval=FALSE}
MBT_ebird<-read_csv("Data/MBT_ebird.csv") #read in my ebird csv file from my github account
```

You can often  read files in directly from the web by putting in the web address as the path.

```{r eval=FALSE}
MBT_ebird<-read_csv("https://github.com/mbtoomey/Biol_7263/blob/main/Data/MBT_ebird.csv?raw=true") #read in my ebird csv file from my github account
```

Readr has a varitey of other functions to import other data formats

- `read_tsv()` - tab separated data
- `read_table()` - fixed width files with white space separation
- `read_delim()` - any delimiter that you specify

### Column Names

By default `read_csv` willl use the first lines as the column names in the resulting tibble. If there are not column names you can omit them:

`MBT_ebird<-read_csv("Data/MBT_ebird.csv", col_names = FALSE)`

or you can provide a vector of names: 

`MBT_ebird<-read_csv("Data/MBT_ebird.csv", co_names = c("entry", "list_ID", "common_name")`

### Parsing

View the imported file

```{r eval=FALSE}
glimpse(MBT_ebird)
```

You will notice that `date` and `time` columns were automatically assigned to the correct data type. Readr uses the following rules, by default to classify data as it is reads in: 

- logical: contains only “F”, “T”, “FALSE”, or “TRUE”.
- integer: contains only numeric characters (and -).
- double: contains only valid doubles (including numbers like 4.5e-5).
- number: contains valid doubles with the grouping mark inside.
- time: matches the default time_format.
- date: matches the default date_format.
- date-time: any ISO8601 date.

If none of these rules apply, then the column will stay as a vector of strings.

For specific data import problems these rules can be overridden with specifications within the the function call. More details here: https://readr.tidyverse.org/articles/readr.html

### Importing directly from Excel and googlesheets **not recommended**

There are tools to do this, but I would avoid this if possible as these are likely to be buggy and unpredictable. 

- Excel - https://readxl.tidyverse.org/
- GoogleSheets - https://googlesheets4.tidyverse.org/

### Exporting data 

Readr can export tibbles you create in R is several formats:

- `write_delim(x, file, delim = " ")` - Write files with any delimiter. 
- `write_csv(x, file)` - Write a comma delimited file. 
- `write_tsv(x, file)` - Write a tab delimited file.

